---
title: "Multivariate Regression Analysis of mtcars Dataset"
author: "Kevin E. D'Elia"
date: "June 24, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Executive Summary

Fuel economy is arguably the most important criteria for car owners.  The purpose of this project is to use multivariate linear regression to analyze the **mtcars** dataset in an effort to understand the relationship between a set of independent predictor variables (_regressors_) and miles per gallon (_response_).  A brief exploratory data analysis will be performed, followed by model selection and then coefficient interpretation.  The analysis will conclude by providing answers to the following two questions:

   * Is an automatic or manual transmission better for MPG?
   * What is the quantitative difference in MPG between automatic and manual transmissions?
   
# Exploratory Data Analysis

The **mtcars** dataset, available in the default R environment, is a `r nrow(mtcars)` by `r ncol(mtcars)` matrix, information about which can be obtained via _?mtcars_.  The `r ncol(mtcars)` variables are: **`r names(mtcars)`**.  Since the analysis is looking at MPG differences between the two transmission types, it makes sense to split the dataset into one for automatic and another for manual transmission.

```{r split.dataset}
split.data <- split(mtcars, as.factor(mtcars$am))
auto.trans <- split.data$`0`
manual.trans <- split.data$`1`
```

The range for MPG in cars with automatic transmission is **`r range(auto.trans$mpg)`** and for manual it is **`r range(manual.trans$mpg)`**, so it would appear that manual transmission offers better fuel economy.  Scatterplots in **_Figure 1_** show a downward trend for MPG against a single regressor for **both** automatic and manual transmissions.  For automatic transmission, the linear line of best fit appears to have a slightly lower slope, but the loess line is more skewed.  Applying linear regression techniques will help confirm this preliminary interpretation.

# Model Selection and Fitting

Choosing a model depends heavily on selecting the most meaningful variables for analysis to avoid over- or under-fitting the model.  As a first step, simple linear regression will be applied, using **wt** (_weight_) as the regressor, since vehicular weight is a key contributor to fuel consumption:  **_lm(mpg ~ wt, data = auto.trans)_** and **_lm(mpg ~ wt, data = manual.trans)_**.  The formulae used result in a model of the form **Y = $\beta_0$ + $\beta_1$X**.

```{r slr, echo=FALSE}
auto.fit.wt <- lm(formula = mpg ~ wt, data = auto.trans)
manual.fit.wt <- lm(formula = mpg ~ wt, data = manual.trans)
```

Of the remaining variables, the next best choice would be **hp** (_horsepower_).  The formula is modified thusly for both datasets: **_mpg ~ wt + hp_**, resulting in a first-order model of the form **Y = $\beta_0$ + $\beta_1X_1$ + $\beta_2X_2$**.

Lastly, an interaction formula is used:  **_mpg ~ wt:hp_**, resulting in a first-order interaction model of the form **Y = $\beta_0$ + $\beta_1X_1X_2$**.

```{r mvlr, echo=FALSE}
auto.fit.wt.hp <- lm(formula = mpg ~ wt + hp, data = auto.trans)
manual.fit.wt.hp <- lm(formula = mpg ~ wt + hp, data = manual.trans)
auto.fit.wt_hp <- lm(formula = mpg ~ wt:hp, data = auto.trans)
manual.fit.wt_hp <- lm(formula = mpg ~ wt:hp, data = manual.trans)
```

# Coefficient Interpretation

$\hat{\alpha}$
$\beta_j$ is the regression coefficient for the _jth_ predictor (slope representing the change in Y for a unit change in X)

# Residuals and Diagnostics
# Conclusions
   Did the student answer the questions of interest or detail why the question(s) is (are) not answerable? 
   “Is an automatic or manual transmission better for MPG”
    "Quantify the MPG difference between automatic and manual transmissions" 
    Did the student quantify the uncertainty in their conclusions and/or perform an inference correctly?
    
    
    
    
    
    
    
    
    
# Appendix of Tables and Figures
```{r eda_auto, echo=FALSE}
library(car)
scatterplot(mpg ~ wt, data = auto.trans, spread = FALSE, smoother.args=list(lty=2), pch=19, main="Automatic", xlab="Wt", ylab = "mgp", cex.main = .9, font.main = 3, cex.axis = .6, cex.lab = .7)
```

***

```{r eda_manual, echo=FALSE}
library(car)
scatterplot(mpg ~ wt, data = manual.trans, spread = FALSE, smoother.args=list(lty=2), pch=19, main="Manual", xlab="Wt", ylab = "mgp", cex.main = .9, font.main = 3, cex.axis = .6, cex.lab = .7)
```

**_Figure 1_**: MPG comparison for automatic and manual transmissions, conditioned by weight

```{r coefficients, echo=FALSE}
coefficients(auto.fit.wt)
coefficients(manual.fit.wt)
coefficients(auto.fit.wt.hp)
coefficients(manual.fit.wt.hp)
coefficients(auto.fit.wt_hp)
coefficients(manual.fit.wt_hp)
```

Did the student do a residual plot and some diagnostics?
Appendix can be no more than three pages